---
title: Basic R functions
author: Hauke Licht
institute: University of Cologne
date: last-modified
date-format: "MMMM D, YYYY"
format: 
  revealjs:
    fontsize: 20pt
    standalone: true
    embed-resources: true
echo: true
code-annotations: select
---

# Introduction

# Recap: Types of vectors
---

## Terminology

- **vector**: a sequence of values
- **element**/**value**: a single value in a vector

```{r}
#| eval: false
vector <- c(1, 2, 3)
value <- vector[1]
```

## Vector Types in R

In R, there are four main types of vectors: 

1. `logical` for "boolean" (true/false) values
2. `integer` for whole numbers (e.g., ... -99, ... -1, 0, 1, 2, ...) 
3. `double` for real-valued numbers (e.g., 1.3)
4. `character` for representing **text**

Each type has its own characteristics and is used for different purposes.

---

## `NA` values

::: {.callout-note icon=false}
`NA` is the special value used in R to represent missing or undefined values
:::


There are special `NA` values for each vector type:

1. `logical` &rarr; `NA`
2. `integer` &rarr; `NA_integer_`
3. `double` &rarr; `NA_real_` (don't ask me why it's not `NA_double_` 😅)
4. `character` &rarr; `NA_character_`

```{r}
typeof(NA)
typeof(as.character(NA))
typeof(NA_character_)
```

---

## Logical Vectors

- logical vectors are used to represent *boolean* values
- boolean values can be either `TRUE` or `FALSE`
- they are commonly used for logical operations and conditional statements.

#### Example

```{r}
x <- c(TRUE, FALSE, TRUE)
typeof(x)                       # <1>
x
```

1. `typeof()` returns the type of the vector

---

## Logical Vectors

```{r}
is.logical(TRUE)
is.logical(FALSE)
is.logical(NA)
```

---

## Integer Vectors

- integer vectors are used to represent whole numbers.
- they are commonly used for counting, indexing, and arithmetic operations.

#### Example

:::: columns

::: column

**_Important!_** Just typing numbers gives you "double" vectors (see next slide)

```{r}
x <- c(1, 2, 3)
typeof(x)
x
```


:::

::: column

Need to convert to integer vector ...

```{r}
x <- as.integer(x)
typeof(x)
x
```

... or use `L` to declare integer explicitly

```{r}
x <- c(1L, 2L, 3L)
typeof(x)
x
```

:::
::::

---

## Functions for creating integer sequences

### seq()

`seq()` generates a sequence of numbers

#### Arguments

- `from`: Starting value of the sequence
- `to`: Ending value of the sequence
- `by`: Step size (default: 1)
- `length.out`: Length of the sequence (alternative to `to`)
- `along.with`: Vector to match the length of the sequence with

#### Example

```{r}
x <- seq(from = 1, to = 10, by = 2)
x
```

---

## Functions for creating integer sequences

### seq_len()

`seq_len()` generates a sequence of numbers from 1 to a specified length

#### Arguments

- `length.out`: Length of the sequence

#### Example

```{r}
x <- seq_len(length.out = 5)
x
```

---

## Functions for creating integer sequences

### seq_along()

`seq_along()` generates a sequence of numbers from 1 to the length of a vector

#### Arguments

- `along.with`: Vector to match the length of the sequence with

#### Example:
```{r}
x <- c("a", "b", "c")
y <- seq_along(along.with = x)
y
```

---

## Double Vectors

- Double vectors are used to represent decimal numbers.
- They are commonly used for mathematical calculations and statistical analysis.

#### Example

```{r}
x <- c(1.5, 2.7, 3.9)
x
```

---

## Character Vectors

- Character vectors are used to represent text or strings.
- They are commonly used for storing and manipulating **textual data**.

#### Example

```{r}
x <- "Hello"
typeof(x)
x
```

# Basic R functions for text wrangling

## Overview

1. Functions for *creating* character vectors
2. Functions for *manipulating* ("changing") character vectors
3. Functions for *analyzing* character vectors

---

## Functions for creating character vectors

### c()

`c()` combines values into a vector

#### Example

```{r}
x <- c("Hello", "World")
x
```

---

## Functions for creating character vectors

### rep()

`rep()` replicates values in a vector

#### Arguments

- `x`: Value to be replicated
- `times`: Number of times to replicate the value
- `each`: Number of times to repeat each value

#### Examples

```{r}
x <- rep("Hello", times = 3)
x
```

```{r}
rep(c("Hello", "World"), each = 2)
rep(c("Hello", "World"), times = 2)
```

---

## Functions for creating character vectors

### paste()

`paste()` combines or concatenates several character values into a single character value

#### Arguments

- `...`: Character vectors to be combined
- `sep`: Separator between the values (default: " ")
- `collapse`: Separator between the combined values (default: NULL)

#### Example

```{r}
x <- c("a", "b", "c")
y <- c("1", "2", "3")
paste(x, y, sep = ":")
```


```{r}
x <- c("a", "b", "c")
paste(x, collapse = ", ")
```

---

## Functions for creating character vectors

### sprintf()

`sprintf()` formats character values according to a specified format

#### Arguments

- `fmt`: Format string
- `...`: Values to be formatted

#### Example

```{r}
sprintf(fmt = "Hello, %s! My name is %s", "friends", "Hauke")
```

*Note:* we use `%s` as a placeholder to insert a character value

## Functions for creating character vectors

### sprintf()

the most important formatting options (a.k.a "placeholders") are:

- `%s` for inserting a character value
- `%d` for inserting an integer value
- `%f` for inserting a double value

See https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/sprintf for more options


## Functions for creating character vectors

### sprintf()

`sprintf()` is great for creating identifiers (IDs):

```{r}
df <- data.frame(
  document_number = c(1, 1, 2, 2),
  paragraph_number = c(1, 2, 1, 2)
)
df$paragraph_id <- sprintf("d%02d_p%02d", df$document_number, df$paragraph_number)
df
```

*Note:* `%02d` is a placeholder for a two-digit integer with leading zeros

```{r}
sprintf("%02d", 8:12)
sprintf("%03d", c(1, 10, 100))
```

---

## Functions for creating character vectors

### Exercise

1. read the sample of the ParlSpeech2 *UK Hose of Commons* corpus I have created
2. group the data by *party* and *date* using `dplyr::group_by`
3. use `dplyr::summarize` and `paste` to aggregate ("combine") speeches by party and date
4. use `dplyr::mutate` and `sprintf` to create a new column with a unique identifier for each party--date text unit

```{r} 
#| eval: false
# step 1
library(readr)
fp <- file.path("data", "datasets", "parlspeech2_gbr_sample.tsv")
df <- read_tsv(fp)
```

---

## Functions for analyzing character vectors

### nchar()

`nchar()` counts the number of characters in a character value

#### Arguments

- `x`: Character vector

#### Example

```{r}
x <- "Hello, World!"
nchar(x)
```

## Functions for analyzing character vectors

### grepl()

`grepl()` tests if a pattern is present in a character vector

#### Arguments

- `pattern`: Pattern to be matched
- `x`: Character vector

#### Example

```{r}
x <- "Hello, World!"
grepl("Hello", x)
grepl("banana", x)
grepl("hello", x, ignore.case = TRUE)
```

*Note:* be careful with `NA` values:

```{r}
grepl(pattern = "Hello", x = NA)
grepl(pattern = NA, x = "Hello")
```

---

## Functions for analyzing character vectors

### grep()

`grep()` returns the indices of the values in a character vector that contain a pattern

#### Arguments

- `pattern`: Pattern to be matched
- `x`: Character vector
- `value`: Logical value (`TRUE`/`FALSE`) indicating whether to return the values instead of the indices

#### Example

```{r}

x <- c("apple", "banana", "cherry")
grep("a", x)
grep("a", x, value = TRUE)
```

---

## Functions for analyzing character vectors

### Exercise (*continued*)

1. Take the data frame of party--date level speech texts created previously
2. Use `dplyr::mutate` and `nchar` to create a new column with the number of characters in each text
3. Use `dplyr::mutate` and `grepl` to create an indicator "mentions_brexit" that is `True` if a party--date text unit contains the term "Brexit" and `False` otherwise
4. Summarize the proportion of party--date text units that contain the term "Brexit" by party and year
5. Analyze:
    1. In what year was the term "Brexit" most prevalent?
    2. Does the answer to 5.1 depend on the party?

---

## Functions for manipulating character vectors

### `trimws`

`trimws()` removes leading and trailing whitespace from a character value

#### Arguments

- `x`: Character vector

#### Example

```{r}
x <- "  Hello, World!  "
trimws(x)
```
---

## Functions for manipulating character vectors

### `tolower()` and `toupper()`

`tolower()` and `toupper()` convert text to lowercase and uppercase, respectively

#### Arguments

- `x`: Character vector

#### Example

```{r}
x <- "Hello, World!"
tolower(x)
toupper(x)
```

---

## Functions for manipulating character vectors

### `substr()`

`substr()` extracts a substring from a character value

#### Arguments

- `x`: Character vector
- `start`: Starting position of the substring
- `stop`: Ending position of the substring

#### Example

```{r}
x <- "Hello, World!"
substr(x, start = 1, stop = 5)
```

---

## Functions for manipulating character vectors

### `sub` and `gsub`

- `sub()` replaces the first occurrence of a pattern in a character value
- `gsub()` replaces all occurrences of a pattern in a character value

#### Arguments

- `pattern`: Pattern to be replaced
- `replacement`: Replacement value
- `x`: Character vector

#### Example

```{r}
x <- "Hello, World!"
sub(pattern = "World", replacement = "Universe", x)
gsub(pattern = "o", replacement = "🤗", x)
```

---

## Functions for manipulating character vectors

### strsplit

`strsplit()` splits a character value into substrings based on a specified delimiter

#### Arguments

- `x`: Character vector
- `split`: Delimiter to split the character value

#### Example

```{r}
x <- "Hello, World!"
strsplit(x, split = ", ")
```

**_Important:_** `strsplit()` returns a **list** of character vectors

---

## Functions for manipulating character vectors

### Exercise (*continued*)

1. Take the data frame of party--date level speech texts with the "mentions_brexit" created previously
2. locate the character positions where the term "Brexit" occurs
3. *advanced*
    1. for each occurrence, extract the term "Brexit" ± 20 characters left and right of it
    2. what are the 20 terms that most freuquently co-occur in the ±20 character window with the term "Brexit"?

---


# `stringr`

## The `stringr` package

- The `stringr` package provides a set of functions for working with character vectors
- The functions in `stringr` are designed to be more consistent and easier to use than the base R functions

#### How to

```{r}
library(stringr)
```

*Note:* `stringr` is part of the "tidyverse" and is loaded automatically when you load the `tidyverse` package

---

## `stringr` equivalents to base R functions

- `str_c()` is equivalent to `paste()`
- `str_length()` is equivalent to `nchar()`
- `str_to_lower()` and `str_to_upper()` are equivalent to `tolower()` and `toupper()`
- `str_sub()` is equivalent to `substr()`
- `str_replace()` and `str_replace_all()` are equivalent to `sub()` and `gsub()`
- `str_split()` is equivalent to `strsplit()`
- `str_detect()` is equivalent to `grepl()`
- `str_locate()` is equivalent to `grep()`

---

## `stringr` equivalents to base R functions

**Practical advantage:** `stringr` functions are defined and named more consistently

- the first argument is always the character vector
- if applicable, the second argument is always the pattern
- if applicable, the third argument is always the replacement value

#### Examples

```{r}
str_replace("Hello, World!", pattern = "World", replacement = "Universe")
# instead of `sub(World, "Universe", "Hello, World!")`

str_detect("Hello, World!", "Hello")
# instead of `grepl("Hello", "Hello, World!")`

str_locate("Hello, World!", "World")
# instead of `grep("World", "Hello, World!")`
```

---

## Other use stringr functions

- `str_trim()`: removes leading and trailing whitespace
- `str_pad()`: pads a string with spaces
- `str_wrap()`: wraps a string to a specified width
- `str_sort()`: sorts a character vector
- `str_order()`: returns the order of a character vector
- `str_replace_na()`: replaces `NA` values with a specified value
- `str_extract()`: extracts a pattern from a character vector

---